Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-03-25 PandaOverview.tex
Iteration III

\mtnote{TODO: replace `resource' with `site'; `http' to `HTTP'?}

The design of the PanDA WMS started in 2005 and its implementation went into
production for the LHC Run 1 on 2009. PanDA was then partially redesigned to be
deployed on Run 2, on 2015. In this section, we summarize the design,
architecture, and execution process of the current state of the practice.

@@

% Consistently, pilots are considered to be multitenant as jobs belonging to
% multiple workflows and/or users can be bound and executed on them.

PanDA' data model uses events as the unit of data\mtnote{Is event an abstraction
specific to the MD workflow?}. Each event represents a collision captured in the
detector and one or more events can be packaged into files or data source with
other container abstractions. As with jobs, data have both attributes and
states, and some of the attributes are shared between events and jobs so to
create associations. Raw, reconstruction, and simulation data are assumed to be
distributed across multiple storage facilities and managed by the ATLAS
Distributed Data Management (DDM)~\cite{garonne2012atlas}. Datasets required by
each job are assumed to be replicated when necessary over the network, both for
input and output data.

@@

The implementation of PanDA consists of several interconnected subsystems, most
of them built from off-the-shelf and Open Source components. At a high level,
PanDA's architecture has five main subsystems: PanDA Server, AutoPyFactory,
PanDA Pilot, PanDA Monitoring, and JEDI. Subsystems communicate via dedicated
API or http messaging. Databases are used to store eventful entities like tasks,
jobs and events, and to store information about resources, logs, and accounting.
Each subsystems is implemented by one or more modules as shown in
Figure~\ref{fig:architecture}.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{figures/panda_architecture.pdf}
  \end{center}
\caption{PanDA architecture representing the subsystems and components relevant
to the integration of PanDA with Titan. Numbers indicates the execution process
based on JEDI, from submission of tasks (1) to retrieval of their output (20).
The monitoring subsystem, the architectural details of PanDA Pilot and the
communication among subsystem's components are abstracted to improve clarity.
% \mtnote{This diagram is a placeholder. I am working on a substitute
% that will be referenced when describing PanDA's subsystems.}Schematic view of
% the PanDA system, as deployed for production in 2017. Black boxes indicate
% PanDA's subsystems; Orange boxes components of a subsystem; purple boxes
% databases; green arrows http messaging interfaces; red arrows API; purple
% arrows database communications.
}
% Originally PanDA was designed for grid infrastructure... In this
% paper is focussed on the HPC use case; there are differences from
% the grid use case. This paper discusses how PanDA has been adapted
% to execute a workload % on HPC resources. The schematic overview is
% presented for workload X on Titan......
\label{fig:architecture}
\end{figure}

\paragraph{\textbf{PanDA Server}} this subsystem is the central hub service of
PanDA and it operates as a web service~\cite{maeno2011overview}. It runs on
Apache web server, interacting with the back-end database running on separate
servers. It implements the Apache worker model, with many independent processes
handling client requests in parallel. Since all state is maintained in the
central database, the PanDA Server application instance itself is stateless.
Currently, the production version of PanDA utilizes an Oracle database as
backend, but PanDA can also work with MySQL family of databases.

@@

PanDA Server consists of four components implemented in Python:
% and communicating via Python APIs
Task Buffer, Brokerage, Job Dispatcher, and Data Service. Task Buffer manages
PanDA's global queue by: (i) storing jobs into the global queue; (ii) providing
information about and updating the state of each stored job. The Brokerage
component: (i) matches job attributes with site and pilot attributes; (ii)
manages the dispatch of input data to processing sites; and (iii) derives
PanDA's data pre-placement requirement. The Job Dispatcher: (i) receives
requests from active pilots for new jobs to process; (ii) retrieves from Task
Buffer the jobs with highest priority and that can be executed on the site of
the requesting pilot; (iii) creates a wrapper script for the selected job,
tailored to the resource of the requesting pilot; and (iv) monitors the state of
the job execution on the pilot, updating the Task Buffer component. Finally, the
Data Service component: (i) retrieves information from Task Buffer and Brokerage
about the data requirements of each job and the site to which each job has been
assigned; (ii) interacts with the ATLAS DDM to make input files available to the
jobs and manage their output data.

@@

\paragraph{\textbf{PanDA Pilot}} this subsystem implements the execution
environment for PanDA's jobs\cite{nilsson2011atlas}. Implemented in Python, a
PanDA Pilot is launched by a wrapper script submitted to a site. Wrapper scripts
are tailored to the middleware exposed by each site so to maintain PanDA Pilot
site-independent. PanDA Pilot: (i) prepares the execution environment for
PanDA's jobs,
% including  recovering and cleaning up previously failed jobs on that site and
collecting information about the compute, data, and network capabilities of the
site; (ii) requests a job for execution to the Job Dispatcher component of PanDA
Server; (iii) initiates and monitors the execution of pulled jobs, including
setting up the job, transferring its input when required, executing the payload,
transferring output, and checking log files and local space; and (iv) cleans up
when the payload has finished to execute.

@@

% -----------------------------------------------------------------------------
% PILOT DESCRIPTION: COMMENTED OUT BECAUSE WE DESCRIBE IT LATER ON.
% PanDA Pilot implements the pilot paradigm~\cite{turilli2015comprehensive},
% decoupling resource acquisition from job execution via multi-stage scheduling.
% This has several benefits, including increasing throughput by lowering the
% time taken to queue and schedule jobs; enabling both concurrent and
% sequential job execution depending on the number of cores available to the
% pilot and the time for which it remains available; and exposing a unified and
% consistent interface for the scheduling of jobs, independent from the
% middleware of the site on which the pilot has been instantiated.
% -----------------------------------------------------------------------------

@@

\paragraph{\textbf{PanDA Monitoring}} this subsystem is implemented as a
web-based, dashboard-style graphical application~\cite{klimentov2011atlas}. It
runs on Apache and interacts with the back-end database of PanDA Server to
enable persistence. PanDA Monitoring enables users and site administrators to
gather information about the status of current and past jobs, data movement, and
pilot factory. PanDA Monitoring allows users access to all log files of each
job, greatly simplifying code debugging and failure analysis in a distributed
computing environment.

@@

The relation between tasks and jobs can be one-to-one or one-to-many, and the
conversion between the two can by static or dynamic. During the LHC Run 1, PanDA
required users or the application layer to perform a static conversion between
tasks and jobs: tasks were described as a set of jobs with a fixed number of
events and then submitted to the PanDA Server (Fig.~\ref{fig:architecture}:A1).

This approach introduced inefficiency both with usability and resource
utilization~\cite{borodin2015unified}. Ideally, users should not have to reason
in terms of jobs: Users conceive analyses in terms of one or more, possibly
related tasks; the `job' abstraction is required exclusively by the execution
middleware, i.e. PanDA. Further, a static partitioning of tasks into jobs does
not take into account the heterogeneity and dynamicity of the resources of the
pilots on which each job will be executed.

@@

JEDI was deployed for the LHC Run 2 to address these inefficiencies with a
machine-driven conversion of tasks into jobs. Users or the application layer (i)
submits tasks descriptions to JEDI; (ii) JEDI partitions tasks to jobs of
different size, depending on both static and dynamic information about available
resources; (iii) jobs are bound to sites with resources that best match jobs'
requirements; and (iv) jobs are submitted to the PanDA Server for execution
(Fig.~\ref{fig:architecture}:A2-B).

Once submitted to the PanDA Server, jobs are stored by the Task Buffer component
into a global queue implemented as a relational database
(Fig.~\ref{fig:architecture}:C). When jobs are submitted directly to the PanDA
Server, the Brokerage component is used to bind jobs to available sites,
depending on static information about the resources available for each site.
Jobs submitted by JEDI are instead already bound to resources so no further
brokerage is needed. Once jobs are bound to sites, the Brokerage module
communicates to the Data Service module what data sets need to be made available
on what site. The Data Service communicates these requirements to the ATLAS DDM
that, when needed, takes care of aggregating files into datasets and containers,
replicating them on the required sites.

The AutoPyFactory subsystem communicates with the Task Buffer component of the
PanDA Server, acquiring information about jobs that have been bound and are
ready for execution. On the base of this information, AutoPyFactory defines a
set of suitable PanDA Pilots, queues them into APFQueues, and concurrently
submits them to multiple sites.

@@

When PanDA Pilots become available on sites, they
% bootstrap, download their code, collect information about the execution
% environment and
call the Job Dispatcher module of the PanDA Server, requesting for a job to
execute \mtnote{Can a pilot request multiple jobs to execute
concurrently?}\sergeynote{In principle it should be able to do that, but that's
not how it is operated. One pilot - one job is a cleaner model in an environment
where resources are "owned" by the organization that runs pilots. Simplifies
many things. At least as far as I know.}. The Job Dispatcher communicates with
the Task Buffer, requesting for a job that is bound to the site of that pilot
and ready to be executed. Task Buffer checks the global queue (i.e., the PanDA
database) and if such a job is available, it passes the job handler to the Job
Dispatcher. The Job Dispatcher dispatches the job to the PanDA Pilot.

Upon receiving a job, a PanDA Pilot starts a monitoring process and forks a
subprocess for the execution of the job's payload. The job is setup, input data
are transferred from the designated staging in location, and the job's payload
is executed. Once completed, output is transferred to the staging out location.
% During the job's payload execution, the monitoring process of PanDA Pilot
% checks the status of the job execution and its environment.

The Data Service module of the PanDA Server tracks and collects the output
generated by each job, updating jobs' attributes via the Task Buffer module.
When the output of all the jobs of a task are retrieved, they are aggregated and
made available to the user either via the PanDA Server of the JEDI service,
depending on how tasks or jobs were initially submitted.

\mtnote{Think whether a comparison with other WMS based on the review we did
makes sense in this context.}\sergeynote{You provided a nice description of the
PanDA system here.}

>>>>

The design of PanDA WMS started in 2005 and its implementation went into
production for the LHC Run 1 on 2009. PanDA was then extended with new
subsystems to be deployed on Run 2, on 2015. In this section, we summarize the
design, architecture, and execution process of the current state of the
practice. We focus on those subsystems and components that are most relevant to
the the integration of PanDA with supercomputers, the main topic of this paper.

% \mtnote{TODO: replace `resource' with `site'; `http' to `HTTP'?}

@@

% Consistently, pilots are considered to be multitenant as jobs belonging to
% multiple workflows and/or users can be bound and executed on them.
% PanDA' data model mostly uses events as the unit of data
% \mtnote{Is event an abstraction specific to the MD workflow?}.

In PanDA's data model, each event represents a collision captured in the
detector and one or more events can be packaged into files or data source with
other container abstractions. As with jobs, data have both attributes and
states, and some of the attributes are shared between events and jobs.
% so to create associations.
Raw, reconstruction, and simulation data are assumed to be distributed across
multiple storage facilities and managed by the ATLAS Distributed Data Management
(DDM)~\cite{garonne2012atlas}. Datasets required by each job are assumed to be
replicated when necessary over the network, both for input and output data.

@@

The implementation of PanDA consists of several interconnected subsystems, most
of them built from off-the-shelf and Open Source components. Subsystems
communicate via dedicated API or HTTP messaging and each subsystems is
implemented by one or more modules. Databases are used to store eventful
entities like tasks, jobs and events, and to store information about sites,
resources, logs, and accounting.

The execution of specific tasks of ATLAS workflows on supercomputers involves
four main subsystems of the PanDA architecture: PanDA Server, AutoPyFactory,
PanDA Pilot, and JEDI. Figure~\ref{fig:architecture} shows a diagrammatic
representation of these subsystems' architecture. For a full list of subsystems
see Ref.~\cite{panda-wiki_url}.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{figures/panda_architecture.pdf}
  \end{center}
    \caption{PanDA architecture with the subsystems and components
    relevant to the integration of PanDA with supercomputers. Numbers indicates
    the execution process based on JEDI: From the submission of tasks (1) to the
    retrieval of their output (20). The monitoring subsystem, the architectural
    details of PanDA Pilot and the communication among subsystem's components
    are abstracted to improve clarity.}
\label{fig:architecture}
\end{figure}

% Originally PanDA was designed for grid infrastructure... In this
% paper is focussed on the HPC use case; there are differences from
% the grid use case. This paper discusses how PanDA has been adapted
% to execute a workload % on HPC resources. The schematic overview is
% presented for workload X on Titan......

\paragraph{\textbf{PanDA Server}} this subsystem is the central hub service of
PanDA and it operates as a web service~\cite{maeno2011overview}. It runs on
Apache web server, interacting with the back-end database running on separate
machines. PanDA Server implements the Apache worker model, with many independent
processes handling client requests in parallel. Since all state is maintained in
the central database, the PanDA Server instance itself is stateless. Currently,
the production version of PanDA utilizes an Oracle database as backend, but
PanDA Server can also work with the MySQL family of databases.

@@

% and communicating via Python APIs

PanDA Server consists of four main components implemented in Python: Task
Buffer, Brokerage, Job Dispatcher, and Data Service. Task Buffer manages PanDA's
global queue by: (i) storing jobs into the global queue; (ii) providing
information about and updating the state of each stored job. The Brokerage
component: (i) matches job attributes with site and pilot attributes; (ii)
manages the dispatch of input data to processing sites; and (iii) derives
PanDA's data pre-placement requirement. The Job Dispatcher: (i) receives
requests from active pilots for new jobs to process; (ii) retrieves from Task
Buffer the jobs with highest priority and that can be executed on the site of
the requesting pilot; (iii) creates a wrapper script for the selected job,
tailored to the site of the requesting pilot; and (iv) monitors the state of
the job execution on the pilot, updating the Task Buffer component. Finally, the
Data Service component: (i) retrieves information from Task Buffer and Brokerage
about the data requirements of each job and the site to which each job has been
assigned; (ii) interacts with the ATLAS DDM to make input files available to the
jobs and manage their output data.

@@

% including  recovering and cleaning up previously failed jobs on that site and

\paragraph{\textbf{PanDA Pilot}} this subsystem implements the execution
environment for PanDA's jobs\cite{nilsson2011atlas}. Implemented in Python, a
PanDA Pilot is launched by a wrapper script submitted to a site. Wrapper scripts
are tailored to the middleware exposed by each site so to maintain PanDA Pilot
site-independent. PanDA Pilot: (i) prepares the execution environment for
PanDA's jobs, collecting information about the compute, data, and network
capabilities of the site; (ii) requests a job for execution to the Job
Dispatcher component of PanDA Server; (iii) initiates and monitors the execution
of pulled jobs, including setting up the job, transferring its input when
required, executing the payload, transferring output, and checking log files and
local space; and (iv) cleans up when the payload has finished to execute.

@@

% -----------------------------------------------------------------------------
% PILOT DESCRIPTION: COMMENTED OUT BECAUSE WE DESCRIBE IT LATER ON.
% -----------------------------------------------------------------------------
% PanDA Pilot implements the pilot paradigm~\cite{turilli2015comprehensive},
% decoupling resource acquisition from job execution via multi-stage scheduling.
% This has several benefits, including increasing throughput by lowering the
% time taken to queue and schedule jobs; enabling both concurrent and
% sequential job execution depending on the number of cores available to the
% pilot and the time for which it remains available; and exposing a unified and
% consistent interface for the scheduling of jobs, independent from the
% middleware of the site on which the pilot has been instantiated.
% -----------------------------------------------------------------------------

@@

% -----------------------------------------------------------------------------
% PANDA MONITORING - SPACE SAVING
% -----------------------------------------------------------------------------
% \paragraph{\textbf{PanDA Monitoring}} this subsystem is implemented as a
% web-based, dashboard-style graphical application~\cite{klimentov2011atlas}. It
% runs on Apache and interacts with the back-end database of PanDA Server to
% enable persistence. PanDA Monitoring enables users and site administrators to
% gather information about the status of current and past jobs, data movement,
% and pilot factory. PanDA Monitoring allows users access to all log files of
% each job, greatly simplifying code debugging and failure analysis in a
% distributed computing environment.
% -----------------------------------------------------------------------------

@@

The relation between tasks and jobs can be one-to-one or one-to-many, and the
conversion between the two can by static or dynamic. During the LHC Run 1, PanDA
required users or applications to perform a static conversion between tasks and
jobs: tasks were described as a set of jobs with a fixed number of events and
then submitted to the PanDA Server (Fig.~\ref{fig:architecture}:A1).

This approach introduced inefficiency both with usability and resource
utilization~\cite{borodin2015unified}. Ideally, users should not have to reason
in terms of jobs: Users conceive analyses in terms of one or more, possibly
related tasks; the `job' abstraction is required by the execution middleware,
i.e. PanDA. Further, a static partitioning of tasks into jobs does not take into
account the heterogeneity and dynamicity of the resources of the pilots on which
each job will be executed.

@@

JEDI was deployed for the LHC Run 2 to address these inefficiencies. Users or
the application layer submits tasks descriptions to JEDI
(Fig.~\ref{fig:architecture}:1) that stores them into a queue implemented by a
database (Fig.~\ref{fig:architecture}:2). Tasks are partitioned into jobs of
different size, depending on both static and dynamic information about available
resources (Fig.~\ref{fig:architecture}:3). Jobs are bound to sites with
resources that best match jobs' requirements, and submitted to the PanDA Server
for execution (Fig.~\ref{fig:architecture}:4).

Once submitted to the PanDA Server, jobs are stored by the Task Buffer component
into a global queue implemented as a  database (Fig.~\ref{fig:architecture}:5).
When jobs are submitted directly to the PanDA Server, the Brokerage component is
used to bind jobs to available sites, depending on static information about the
resources available for each site. Jobs submitted by JEDI are already bound to
sites so no further brokerage is needed. Once jobs are bound to sites, the
Brokerage module communicates to the Data Service module what data sets need to
be made available on what site (Fig.~\ref{fig:architecture}:6). The Data Service
communicates these requirements to the ATLAS DDM (Fig.~\ref{fig:architecture}:7)
that, when needed, takes care of aggregating files into datasets and containers,
replicating them on the required sites (Fig.~\ref{fig:architecture}:8).

The AutoPyFactory subsystem communicates with the Task Buffer component of the
PanDA Server, acquiring information about jobs that are ready for execution on
specific (type of) sites (Fig.~\ref{fig:architecture}:9). On the base of this
information, AutoPyFactory defines a set of suitable PanDA Pilots and
concurrently submits them a Condor-G broker (Fig.~\ref{fig:architecture}:10).
This broker submits these pilots wrapped as jobs or VMs to the required sites
(Fig.~\ref{fig:architecture}:11).

@@

When a PanDA Pilot becomes available on a site, it
% bootstrap, download their code, collect information about the execution
% environment and
calls the Job Dispatcher module of the PanDA Server, requesting for a job to
execute (Fig.~\ref{fig:architecture}:12). The Job Dispatcher communicates with
the Task Buffer, requesting for a job that is bound to the site of that pilot
and ready to be executed. Task Buffer checks the global queue (i.e., the PanDA
database) and if such a job is available, it passes the job handler to the Job
Dispatcher. The Job Dispatcher dispatches the job to the PanDA Pilot
(Fig.~\ref{fig:architecture}:13).

% \mtnote{Can a pilot request multiple jobs to execute
% concurrently?}\sergeynote{In principle it should be able to do that, but
% that's not how it is operated. One pilot - one job is a cleaner model in an
% environment where resources are ``owned'' by the organization that runs
% pilots. Simplifies many things. At least as far as I know.}

Upon receiving a job, a PanDA Pilot starts a monitoring process and forks a
subprocess for the execution of the job's payload. The job is setup, input data
are transferred from the designated staging in location
(Fig.~\ref{fig:architecture}:14), and the job's payload is executed
(Fig.~\ref{fig:architecture}:15). Once completed, output is transferred to the
staging out location (Fig.~\ref{fig:architecture}:16).

% During the job's payload execution, the monitoring process of PanDA Pilot
% checks the status of the job execution and its environment.

The Data Service module of the PanDA Server tracks and collects the output
generated by each job (Fig.~\ref{fig:architecture}:17), updating jobs'
attributes via the Task Buffer module (Fig.~\ref{fig:architecture}:18). When the
output of all the jobs of a task are retrieved, it is made available to the user
via PanDA Server. When a task is submitted to JEDI, task is marked as done
(Fig.~\ref{fig:architecture}:19) and the result of its execution is made
available to the user by JEDI (Fig.~\ref{fig:architecture}:20).

% \mtnote{Think whether a comparison with other WMS based on the review we did
% makes sense in this context.}\sergeynote{You provided a nice description of
% the PanDA system here.}
